---
title: "Project title"
author: "by Team-Name: User1, User2, User3, User4 & User5"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.




```


## Research Question
We are looking at which factors most effect A&E waiting times. We have established 2 main possible factors: 
  1. The impact of the COVID pandemic
  2. The impact of staff Levels
We felt that during the time period for the data set these would be some of the main factors effecting wait times in for A&E. 




## Data
In this study we used three datasets to analyse. 
The first of which was the data set "AandE" this included information on the waiting times for different A&E wards around the country. It included varibles such as number of attendences in a day, and the % of these which were seen to in a certain time period (i.e. %. within 4 hours). 
One of the secondary data sets that were used was "covid" this was another data set produced by public health Scotland. It was used to establish the amount of hospitalisations there were due to COVID in a given week.
The other secondary dataset used was also by Public Health Scotland. This dataset included information about staff vacancies by quarter.


See http://libraryguides.vu.edu.au/c.php?g=386501&p=4347840 for guidance on proper citation for datasets. 
If you got your data off the web, make sure to state the retrieval date.






## Findings


The write-up of your project and findings go here. 

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.





## References

List any references here. You should, at a minimum, list your data source.


