---
title: "Project title"
author: "by Binary_Brigade: Connor Finlay, Ade Obawole, Murray Bone, Alexander Patton & Cameron Smith"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```

```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.




```

## Research Question

We are looking at which factors most effect A&E waiting times.
We have established 2 main possible factors: 1.
The impact of the COVID pandemic 2.
The impact of staff Levels We felt that during the time period for the data set these would be some of the main factors effecting wait times in for A&E.

## Data

In this study we used three datasets to analyse.
The first of which was the data set "AandE" (Public Health Scotland [PHS], 2023a) included information on the waiting times for different A&E wards around the country.
It included varibles such as number of attendences in a day, and the % of these which were seen to in a certain time period (i.e. %. within 4 hours).
One of the secondary data sets that were used was "covid" (PHS, 2023b) this was another data set produced by public health Scotland.
It was used to establish the amount of hospitalisations there were due to COVID in a given week.
The other secondary dataset used was also by Public Health Scotland.
This dataset included information about staff vacancies by quarter (Turas Data Intelligence, 2023).

## Findings

**Methods:**

In our research, we initiated the process by loading three integral datasets - data from A&E, COVID-related data, and information from the NHS workforce.
Our next step was to prime the data for a thorough analysis.
This required us to undertake a series of crucial data cleaning and preprocessing tasks.
Notably, we converted the date columns into the appropriate datetime format and extracted quarters and years to facilitate time-based evaluation, ensuring uniformity across datasets.
Additionally, we conducted a careful selection process to keep only relevant columns, while discarding any that were irrelevant or redundant.

To provide a more comprehensive analysis of A&E data, we further refined our data preprocessing efforts.
We renamed columns for consistency, making it easier to work with the dataset.
Additionally, we separated the HBT (Health Board Trust) and the number of attendances, which allowed us to create categorical variables and convert the number of attendances into a numeric format, essential for calculating measures of central tendency.

We used a variety of data science tools in this investigation.
To learn more about the central tendencies of variables like waiting times, we ran descriptive statistical calculations, such as mean and median calculations.
As we employed a range of plots and graphs to show the trends, patterns, and correlations present in the data, data visualization played a crucial role.
Plots such as scatter plots, bar plots, and time series plots were some examples of these visual aids.

Z-scores allowed us to identify anomalies in the 'PercentageWithin4HoursEpisode' column, highlighting individual data points that significantly deviated from the average waiting time.
A comprehensive time series analysis was conducted, examining changes in waiting times on a quarterly and annual basis.
In order to gain a deeper understanding of these patterns, we created time series graphs that allowed us to observe the variations over time, particularly the changes in the median percentage of patients seen in less than four hours.

Correlation analysis was also used to investigate potential correlations between variables.
Notably, we discovered a negative association between A&E visits within 4 hours and NHS employment vacancies.
To study this link further, a preliminary linear regression model was built.

These integrated data science tools, together with thorough data cleaning and preparation, enabled us to acquire significant insights into waiting times, discover patterns, and investigate potential links within the data, forming a solid foundation for our ongoing investigation.

We had to calculate percentages, mutate our dataframes and create new ones to quantify to mean attendances per Health Board (HBT).
Additionally, we created categorical variables quantifying HBTs into "High", "Medium" "Low" and plotted graphs based on those.

Additionally Renamed columns to maintain consistency as some columns were in different formats.

**Results:**

#### Exploratory Data Analysis (EDA)

During our in-depth analysis of the data, we thoroughly examined crucial variables related to A&E waiting times, COVID hospitalizations, and NHS staff vacancies.
This investigation establishes a solid foundation for gaining a deeper comprehension of the factors that affect A&E waiting periods.

We initiated our analysis by looking into the key summary statistics, which offered vital understanding of our datasets.
Delving into COVID hospitalizations, we uncovered a weekly average of 49.5 hospitalizations during the pandemic, highlighting the time-based variations in these figures.
In a similar fashion, evaluating NHS staff vacancies led us to identify a median of 536 vacancies per quarter, emphasizing the ever-changing staffing levels throughout the monitoring period.

To enhance our understanding of the data, we turned to visualizations.
Histograms and box plots allowed us to visualize the distribution of A&E waiting times, shedding light on variations and the potential presence of outliers.
Furthermore, time series plots provided valuable insights into the trends in A&E waiting times over different time periods.
Notably, these plots revealed fluctuations and a significant decline in waiting times during the COVID pandemic, with a particular focus on the median percentage of patients seen within four hours.
Additionally, we employed Z-scores for anomaly detection, enabling us to pinpoint weeks with unusual waiting time patterns.

#### Correlation Analysis

Our investigation extended to correlation analysis, where we sought to uncover relationships between variables.
Notably, we explored the correlation between A&E visits and staff vacancies.
Our analysis yielded a negative correlation between the median percentage of A&E visits seen within four hours and NHS employment vacancies.
This finding indicates that as staff vacancies increased, the percentage of A&E visits seen within four hours decreased.
The strength and direction of this correlation were visually represented through scatter plots and correlation matrices, further reinforcing the inverse relationship between A&E visits and staff vacancies.

#### Linear Regression Model

To delve deeper into the impact of staff vacancies on A&E waiting times, we developed a preliminary linear regression model.
The model results were enlightening, revealing a statistically significant inverse relationship between the median percentage of A&E visits seen within four hours and NHS employment vacancies.
The R-squared value of 0.90 indicated that over 90% of the variance in the percentage of visits within four hours could be explained by staff vacancies.
This suggests that a higher number of staff vacancies in NHS Scotland is associated with lower performance in meeting the target of seeing A&E patients within four hours.

The combined results offer a thorough summary of our data analysis and its significance in comprehending A&E waiting times, especially during the COVID pandemic and considering NHS staffing levels.
In the following "Discussion" section, we will dive further into our research question responses, examine the strengths and weaknesses of our inquiry, and address the ethical aspects surrounding our conclusions.

**Discussion:**

We discovered that the size of an A&E facility does in fact affect the waiting times for a location/Hea, this was represented by a dataframe that was mutated to demonstrate mean attendances for Health Boards and the percentage of waiting times (within 4 hours and over 4/8/12 hours) for individuals who attended the facility.

There are severe ethical considerations in relation to our investigation as the wait time for an individual within an A&E facility can mean a matter of life and death.

To incorporate our investigation into the Scottish NHS it would be advisable to focus on amount of staff and precautions for if staff are absent as demonstrated by our investigation, size of Health Board and Staff Absences are are significant factor in delayed wait times.

There are limitations of the data source as there are problems with staff vacancies since there is a small sample size which means the data could be distorted.

We Discovered that mathematically there is a relationship between COVID admissions and A&E admissions. The linear regression suggests that whilst there were high levels of COVID hospital admissions the level of A&E admissions fell. At a rate of 2.8 admissions per COVID admission. Additionally the model predicts that at 0 COVID admissions there would be around 24,500 A&E admissions, which positive in regards to the performance of the model as when you take the mean of the admissions data before COVID you get levels roughly akin to this figure. This explains the spike in the % of people waiting for less than four hours during 2020. However it does not explain the decline in this level that followed. One on the limitations of the COVID data set, is how the data was recorded. COVID hospitalizations were registered when someone arrived in hospital and tested positive for the virus, so this can't give us indications of how busy the hospitals were. It mainly tells us how much virus was in the population at the time. 

## References

- Public Health Scotland [PHS]. (2023a). Weekly Accident & Emergency Activity and Waiting Times. https://www.opendata.nhs.scot/dataset/weekly-accident-and-emergency-activity-and-waiting-times/resource/a5f7ca94-c810-41b5-a7c9-25c18d43e5a4. Retrieved 20 October 2023.  
- Public Health Scotland [PHS]. (2023b). Weekly Admissions by Age. https://www.opendata.nhs.scot/ne/dataset/b318bddf-a4dc-4262-971f-0ba329e09b87/resource/b5e3fa11-8a85-4946-bbb2-2e800d4e3594. Retrieved 8 November 2023
- Turas Data Intelligence. [TDI]. (2023). NHS Scotland workforce. https://turasdata.nes.nhs.scot/data-and-reports/official-workforce-statistics/all-official-statistics-publications/05-september-2023-workforce/dashboards/nhs-scotland-workforce/?pageid=9984. Retrieved 13 November 2023. 

